```{r set}
#| include: false
library(tidyverse)
```

# Species {.unnumbered}

Currently only LA, SLA, C, and N values are available and alometric parameters can be derived from inventories (see Data/Forest). Additional data will be needed from TRY [@kattge2019], TALLO [@jucker2022], and @schmitt2024.

### LBA-ECO CD-02 Forest Canopy Structure, Tapajos National Forest, Brazil: 1999-2003

Cite: Ehleringer, J., L.A. Martinelli and J.P. Ometto. 2011. LBA-ECO CD-02 Forest Canopy Structure, Tapajos National Forest, Brazil: 1999-2003. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. doi:10.3334/ORNLDAAC/1009

```{r sp1}
#| message: false
#| warning: false
read_csv("data/raw_data/CD02_Forest_Canopy_Structure_1009/data/CD02_SLA_measurements_TNF.csv", skip = 22, na = "-9999") %>% # nolint
  rename_all(tolower) %>%
  filter(functional_group == "Tree", species != "Not provided") %>%
  select(species, leaf_area, sla) %>%
  gather(trait, value, -species) %>%
  na.omit() %>%
  group_by(species, trait) %>%
  summarise(value = mean(value)) %>%
  mutate(trait = recode(trait,
    "leaf_area" = "LA [m2]",
    "sla" = "SLA [m2/g]"
  )) %>%
  ggplot(aes(value)) +
  geom_density() +
  facet_wrap(~trait, scales = "free") +
  theme_bw() +
  theme(axis.title = element_blank()) +
  ggtitle("N = 54")
```

### LBA-ECO CD-02 Leaf Level Gas Exchange, Chemistry, and Isotopes, Amazonia, Brazil

Cite: Ehleringer, J., L.A. Martinelli, C. Cook, T.F. Domingues, L. Flanagan, J. Berry, and J.P. Ometto. 2011. LBA-ECO CD-02 Leaf Level Gas Exchange, Chemistry, and Isotopes, Amazonia, Brazil. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. doi:10.3334/ORNLDAAC/1010

```{r sp2}
#| message: false
#| warning: false
read_csv("data/raw_data/CD02_Leaf_Level_Gas_Exchange_1010/data/CD02_Leaf_Level_Gas_Exchange.csv", skip = 48, na = "-9999") %>% # nolint
  rename_all(tolower) %>%
  filter(functional_group == "Tree", species != "Not provided") %>%
  select(species, leaf_area, c_conc, n_conc) %>%
  gather(trait, value, -species) %>%
  na.omit() %>%
  group_by(species, trait) %>%
  summarise(value = mean(value)) %>%
  mutate(trait = recode(trait,
    "leaf_area" = "LA [m2]",
    "c_conc" = "C [%]",
    "n_conc" = "N [%]"
  )) %>%
  ggplot(aes(value)) +
  geom_density() +
  facet_wrap(~trait, scales = "free") +
  theme_bw() +
  theme(axis.title = element_blank()) +
  ggtitle("N = 20")
```
