```{r set}
#| include: false
library(tidyverse)
library(knitr)
```

# Climate {.unnumbered}

Climate data are available for both sites (Km67 and Km83) with needed TROLL inputs: rainfall, snet, temperature, wind speed, and vpd (to derive from humidity). The only thing to explore is how to model the effect of Throughfall Exclusion Experiment. There are values and relations given in @nepstad2002 and @nepstad2007, and this should be validated with soil water content (see Data/Fluxes).

## Undisturbed

### LBA-ECO CD-32 LBA Model Intercomparison Project (LBA-MIP) Meteorological Forcing Data

Cite: de Goncalves, L.G.G., N. Restrepo-Coupe, H.R. da Rocha, S.R. Saleska, and R. Stockli. 2013. LBA-ECO CD-32 LBA Model Intercomparison Project (LBA-MIP) Meteorological Forcing Data. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, USA http://dx.doi.org/10.3334/ORNLDAAC/1177

```{r clim}
#| message: false
#| warning: false
list.files("data/raw_data/CD32_LBA_MIP_Drivers_1177/data/Santarem_KM67/",
  pattern = ".txt", full.names = TRUE
) %>%
  read_tsv(
    skip = 16, na = "-9999",
    col_names = c(
      "year", "day", "hour", "minute",
      "tas", "tas_gf", "q", "q_gf", "ws", "ws_gf",
      "pr", "pr_gf", "p", "p_gf", "swdown", "swdown_gf",
      "lwdown", "lwdown_gf", "co2", "co2_gf"
    )
  ) %>%
  mutate(date = as_date(paste0(year, "-01-01")) + day - 1) %>%
  mutate(date = as_datetime(date) + hour * 60 * 60 + minute * 60) %>%
  mutate(date = date - 4 * 60 * 60) %>%
  select(-year, -day, -hour, -minute) %>%
  select(date, tas, q, ws, pr, p, swdown, lwdown, co2) %>%
  gather(variable, value, -date) %>%
  mutate(variable = recode(variable,
    "tas" = "Temperature [K]",
    "q" = "Specific humidity [kg/kg]",
    "ws" = "Wind speed [m/s]",
    "pr" = "Precipitation [kg/m2/s]",
    "p" = "Pressure [Pa]",
    "swdown" = "Shortwave radiation [W/m2]",
    "lwdown" = "Longwave radiation [W/m2]",
    "co2" = "CO2 [ppmv]"
  )) %>%
  ggplot(aes(date, value)) +
  geom_line() +
  facet_wrap(~variable, scales = "free") +
  theme_bw() +
  theme(axis.title = element_blank())
```

### LBA-ECO CD-10 CO2 & H2O Eddy Flux Data at km 67 Tower Site, Tapajos National Forest

Cite: Hutyra, L., S. Wofsy and S. Saleska. 2008. LBA-ECO CD-10 CO2 and H2O Eddy Flux Data at km 67 Tower Site, Tapajos National Forest. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. doi:10.3334/ORNLDAAC/860

```{r eddy}
#| message: false
#| warning: false
read_csv("data/raw_data/CD10_EddyFlux_Tapajos_860/data/km67_eddyflux_2002_2006.txt") %>% #nolint
  rename_all(tolower) %>%
  mutate(date = as_datetime("2000/01/01") + hours * 60 * 60 - 4 * 60 * 60) %>%
  rename(h2o = h2o.mmol.m., p = pamb.pa, lh = lhdry, pr = rain) %>%
  select(
    date, ws, wdir, tamb, tdew, fheat, co2, h2o, p, h, lh, netrad, par, pr,
    nee, r, gee
  ) %>%
  gather(variable, value, -date) %>%
  mutate(variable = recode(variable,
    "tamb" = "Temperature [°C]",
    "tdew" = "Dewpoint [°C]",
    "ws" = "Wind speed [m/s]",
    "wdir" = "Wind direction [°]",
    "fheat" = "Heat flux [°C/m/s]",
    "co2" = "CO2 [mmol/mol]",
    "h2O" = "H2O [mmol/mol]",
    "p" = "Pressure [Pa]",
    "h" = "Sensible heat [W/m2]",
    "lh" = "Latent heat [W/m2]",
    "netrad" = "Net Radiation [W/m2]",
    "par" = "PAR [umol/m2/s]",
    "pr" = "Precipitation [mm]",
    "nee" = "NEE [umol/m2/s]",
    "r" = "Respiration [umol/m2/s]",
    "gee" = "GEE [umol/m2/s]"
  )) %>%
  ggplot(aes(date, value)) +
  geom_line() +
  facet_wrap(~variable, scales = "free") +
  theme_bw() +
  theme(axis.title = element_blank())
```

## TEE

### LBA-ECO CD-05 Soil VWC and Meteorology, Rainfall Exclusion, Tapajos National Forest

Cite: Nepstad, D.C., P.R. Moutinho, and P. Brando. 2013. LBA-ECO CD-05 Soil VWC and Meteorology, Rainfall Exclusion, Tapajos National Forest. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, USA http://dx.doi.org/10.3334/ORNLDAAC/1169

```{r rainfall}
#| message: false
#| warning: false
read_csv("data/raw_data/CD05_Micromet_1169/data/Rainfall_REE_plots_1999_2006.csv", #nolint
  skip = 12, na = "-9999"
) %>%
  rename_all(tolower) %>%
  mutate(date = paste0(
    str_sub(date, 1, 4), "-",
    str_sub(date, 5, 6), "-",
    str_sub(date, 7, 8)
  )) %>%
  mutate(date = as_date(date)) %>%
  gather(site, precipitation, -date, -exclusion) %>%
  mutate(site = recode(site,
    "ppt_casa" = "Casa Onca (~1-km)",
    "ppt_ree_tower" = "Control plot tower"
  )) %>%
  ggplot(aes(date, precipitation, fill = exclusion == 1)) +
  geom_col() +
  theme_bw() +
  facet_wrap(~site, nrow = 2) +
  xlab("") +
  ylab("Daily rainfall [mm]") +
  scale_fill_discrete("Exclusion") +
  theme(legend.position = "bottom")
```

Precipitation data to force the model. ⚠️ not measured at weekend so Monday is three days precipitation. For the moment I have nothing else that @nepstad2002 relation to compute avoided precipitation with precipitation excluded = 0.484 precipitation input + 0.073 as I couldn't get hold of the original data yet. There is also a 60% exclusion reported in @nepstad2007.

## Logged

### LBA-ECO CD-32 LBA Model Intercomparison Project (LBA-MIP) Meteorological Forcing Data

Cite: de Goncalves, L.G.G., N. Restrepo-Coupe, H.R. da Rocha, S.R. Saleska, and R. Stockli. 2013. LBA-ECO CD-32 LBA Model Intercomparison Project (LBA-MIP) Meteorological Forcing Data. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, USA http://dx.doi.org/10.3334/ORNLDAAC/1177

```{r climlogged}
#| message: false
#| warning: false
list.files("data/raw_data/CD32_LBA_MIP_Drivers_1177/data/Santarem_KM83/",
  pattern = ".txt", full.names = TRUE
) %>%
  read_tsv(
    skip = 16, na = "-9999",
    col_names = c(
      "year", "day", "hour", "minute",
      "tas", "tas_gf", "q", "q_gf", "ws", "ws_gf",
      "pr", "pr_gf", "p", "p_gf", "swdown", "swdown_gf",
      "lwdown", "lwdown_gf", "co2", "co2_gf"
    )
  ) %>%
  mutate(date = as_date(paste0(year, "-01-01")) + day - 1) %>%
  mutate(date = as_datetime(date) + hour * 60 * 60 + minute * 60) %>%
  mutate(date = date - 4 * 60 * 60) %>%
  select(-year, -day, -hour, -minute) %>%
  select(date, tas, q, ws, pr, p, swdown, lwdown, co2) %>%
  gather(variable, value, -date) %>%
  mutate(variable = recode(variable,
    "tas" = "Temperature [K]",
    "q" = "Specific humidity [kg/kg]",
    "ws" = "Wind speed [m/s]",
    "pr" = "Precipitation [kg/m2/s]",
    "p" = "Pressure [Pa]",
    "swdown" = "Shortwave radiation [W/m2]",
    "lwdown" = "Longwave radiation [W/m2]",
    "co2" = "CO2 [ppmv]"
  )) %>%
  ggplot(aes(date, value)) +
  geom_line() +
  facet_wrap(~variable, scales = "free") +
  theme_bw() +
  theme(axis.title = element_blank())
```

## 
