```{r set}
#| include: false
library(tidyverse)
```

# Climate {.unnumbered}

## Undisturbed

## TEE

### LBA-ECO CD-05 Soil VWC and Meteorology, Rainfall Exclusion, Tapajos National Forest

Cite: Nepstad, D.C., P.R. Moutinho, and P. Brando. 2013. LBA-ECO CD-05 Soil VWC and Meteorology, Rainfall Exclusion, Tapajos National Forest. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, USA http://dx.doi.org/10.3334/ORNLDAAC/1169

```{r rainfall}
#| message: false
#| warning: false
read_csv("data/raw_data/CD05_Micromet_1169/data/Rainfall_REE_plots_1999_2006.csv",
         skip = 12, na = "-9999") %>% 
  rename_all(tolower) %>% 
  mutate(date = paste0(str_sub(date, 1,4), "-",
                       str_sub(date, 5,6), "-",
                       str_sub(date, 7,8))) %>% 
  mutate(date = as_date(date)) %>% 
  gather(site, precipitation, -date, -exclusion) %>% 
  mutate(site = recode(site, 
                       "ppt_casa" = "Casa Onca (~1-km)",
                       "ppt_ree_tower" = "Control plot tower")) %>% 
  ggplot(aes(date, precipitation, fill = exclusion == 1)) +
  geom_col() +
  theme_bw() +
  facet_wrap(~ site, nrow = 2) +
  xlab("") +
  ylab("Daily rainfall [mm]") +
  scale_fill_discrete("Exclusion") +
  theme(legend.position = "bottom")
```

Precipitation data to force the model.
