```{r set}
#| include: false
library(tidyverse)
```

# Forest {.unnumbered}

We have spatialized repeated inventories with taxonomy for forest dynamics (recruitment, growth, mortality), structure, and composition including control and TEE plots to evaluate and calibrate TROLL simulations. We can also leverage back lidar scanning from @schmitt2024.

## Undisturbed

### LBA-ECO CD-10 TREE DBH MEASUREMENTS AT THE KM 67 TOWER SITE, TAPAJOS NATIONAL FOREST

Cite: Wofsy, S. C., S. R. Saleska, E. H. Pyleand L. R. Hutyra. 2008. LBA-ECO CD-10 Tree DBH Measurements at the km 67 Tower Site, Tapajos National Forest.Data set. Available on-line \[http://www.daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. doi:10.3334/ORNLDAAC/859

```{r inveotry1}
#| message: false
#| warning: false
read_csv("data/raw_data/CD10_DBH_Tapajos_859/data/km_67_tree_surveys_99_05.csv",
  na = "-9999"
) %>%
  rename_all(tolower) %>%
  select(trans, tag, sci_name, fam, x, y, dbh_99, dbh_01, dbh_05, wood_dens) %>%
  gather(year, dbh, -trans, -tag, -sci_name, -fam, -x, -y, -wood_dens) %>%
  mutate(year = as.numeric(recode(year, "dbh_99" = "1999",
                                  "dbh_01" = "2001", "dbh_05" = "2005"))) %>%
  mutate(dbh = as.numeric(dbh)) %>%
  ggplot(aes(x, y, size = dbh, col = wood_dens)) +
  facet_grid(trans ~ year) +
  geom_point() +
  theme_bw() +
  theme(axis.title = element_blank()) +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(0.5, 3))
```

### LBA-ECO CD-10 GROUND-BASED BIOMETRY DATA AT KM 67 TOWER SITE, TAPAJOS NATIONAL FOREST

Cite: Rice, A. H., E. P. Hammond, S. R. Saleska, L. Hutyra, M. Palace, M. Keller,P. B. de Carmargo, K. Portilho, D. Marques and S. C.Wofsy. 2007. LBA-ECO CD-10 Ground-based Biometry Data at km 67 Tower Site, Tapajos National Forest. Data set. Available on-line \[http://www.daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. doi:10.3334/ORNLDAAC/854

```{r inveotry2}
#| message: false
#| warning: false
read_csv("data/raw_data/CD10_Biometry_Tapajos_854/data/lba_km67_dend_dbhs_filled_archive.txt", na = "-9999") %>% #nolint
  rename_all(tolower) %>%
  select(t, tag, family, genus, spp, x, y, starts_with("doysince")) %>%
  gather(date, dbh, -t, -tag, -family, -genus, -spp, -x, -y) %>%
  mutate(date = as.numeric(gsub("doysince1999.", "", date))) %>%
  mutate(date = as_date("1999/1/1") + date) %>%
  mutate(dbh = as.numeric(dbh)) %>%
  ggplot(aes(date, dbh,
    col = paste(family, genus, spp),
    group = paste(t, tag)
  )) +
  geom_line(alpha = 0.2) +
  theme_bw() +
  scale_y_log10() +
  ylab("Diameter at breast height [ cm ]") +
  scale_color_discrete(guide = "none") +
  ggtitle("Spatialised with height")
```

### LBA-ECO CD-15 LAI and Productivity Data, km 67, Tapajos National Forest

Cite: Costa, M.H. and W. Cohen. 2013. LBA-ECO CD-15 LAI and Productivity Data, km 67, Tapajos National Forest: 2003-2004. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. http://dx.doi.org/10.3334/ORNLDAAC/1167

```{r dendrometer}
#| message: false
#| warning: false
read_csv("data/raw_data/CD15_Productivity_1167/data/CD15_Dendrometer_measurements_2003_2004.csv", #nolint
  skip = 24, na = "-9999"
) %>%
  rename_all(tolower) %>%
  gather(
    date, dendrometer,
    -plot, -subplot, -tree_no, -diameter, -comments
  ) %>%
  mutate(date = gsub("band_", "", date)) %>%
  mutate(
    month = str_sub(date, 1, 2),
    year = str_sub(date, 3, 6)
  ) %>%
  mutate(date = as_date(paste0(year, "-", month, "-01"))) %>%
  select(-year, -month) %>%
  ggplot(aes(date, dendrometer,
    group = paste(plot, subplot, tree_no)
  )) +
  geom_line(alpha = .2) +
  theme_bw() +
  xlab("") +
  ylab("Diameter (dendrometer) [ mm ]")
```

Dendrometer data available for fine temporal scale dynamics assessment.

## TEE

### LBA-ECO LC-14 Biophysical Measurements, Rainfall Exclusion, Tapajos National Forest

Cite: Nepstad, D.C. and P.R. Moutinho. 2013. LBA-ECO LC-14 Biophysical Measurements, Rainfall Exclusion, Tapajos National Forest. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, USA http://dx.doi.org/10.3334/ORNLDAAC/1196

```{r dendrometer2}
#| message: false
#| warning: false
read_csv("data/raw_data/LC14_Aboveground_Prod_1196/data/Dendrometer_DBH_REE_Plots_1999_2006.csv", #nolint
  skip = 19, na = "-9999"
) %>%
  rename_all(tolower) %>%
  select(
    treatment, tree_id, life_form, family, genus_species, common_name,
    height_m, starts_with("dbh_"), date_death, field_notes
  ) %>%
  gather(
    date, dbh,
    -treatment, -tree_id, -life_form, -family,
    -genus_species, -common_name,
    -height_m, -dbh_init, -date_death, -field_notes
  ) %>%
  mutate(date = gsub("dbh_", "", date)) %>%
  mutate(
    month = str_sub(date, 1, 3),
    year = str_sub(date, 4, 5)
  ) %>%
  mutate(date = paste0(year, "-", month, "-01")) %>%
  mutate(date = as_date(date)) %>%
  select(-year, -month) %>%
  filter(life_form == "tree") %>%
  ggplot(aes(date, dbh,
    col = genus_species,
    group = paste(treatment, tree_id)
  )) +
  geom_line(alpha = .2) +
  theme_bw() +
  xlab("") +
  facet_wrap(~treatment) +
  ylab("Diameter at breast height [ cm ]") +
  scale_color_discrete(guide = "none") +
  scale_y_log10()
```

Can be used to assess TEE on dynamics simulation in TROLL.

```{r inventory}
#| message: false
#| warning: false
read_csv("data/raw_data/LC14_Aboveground_Prod_1196/data/Stem_inventory_REE_plots_2000_2004.csv", skip = 15, na = "-9999") %>% #nolint
  rename_all(tolower) %>%
  filter(life_form == "TREE") %>%
  ggplot(aes(dbh_2000, fill = obs_final)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~treatment, nrow = 2) +
  scale_x_log10() +
  scale_y_log10() +
  ylab("") +
  xlab("DBH 2000 [ cm ]") +
  scale_fill_discrete("")
```

To assess treatment effect on mortality and growth (DBH 2004 available).
