```{r set}
#| include: false
library(tidyverse)
```

# Forest {.unnumbered}

## Undisturbed

### LBA-ECO CD-15 LAI and Productivity Data, km 67, Tapajos National Forest

Cite: Costa, M.H. and W. Cohen. 2013. LBA-ECO CD-15 LAI and Productivity Data, km 67, Tapajos National Forest: 2003-2004. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. http://dx.doi.org/10.3334/ORNLDAAC/1167

```{r dendrometer}
#| message: false
#| warning: false
read_csv("data/raw_data/CD15_Productivity_1167/data/CD15_Dendrometer_measurements_2003_2004.csv", 
         skip = 24, na = "-9999") %>% 
  rename_all(tolower) %>% 
  gather(date, dendrometer,
         -plot, -subplot, -tree_no, -diameter, -comments) %>% 
  mutate(date = gsub("band_", "", date)) %>% 
  mutate(month = str_sub(date,1,2),
         year = str_sub(date,3,6)) %>% 
  mutate(date = as_date(paste0(year, "-", month, "-01"))) %>% 
  select(-year, -month) %>% 
  ggplot(aes(date, dendrometer, 
             group = paste(plot, subplot, tree_no))) +
  geom_line(alpha = .2) +
  theme_bw() +
  xlab("") +
  ylab("Diameter (dendrometer) [ mm ]")
```

Dendrometer data available for fine temporal scale dynamics assessment.

## TEE

### LBA-ECO LC-14 Biophysical Measurements, Rainfall Exclusion, Tapajos National Forest

Cite: Nepstad, D.C. and P.R. Moutinho. 2013. LBA-ECO LC-14 Biophysical Measurements, Rainfall Exclusion, Tapajos National Forest. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, USA http://dx.doi.org/10.3334/ORNLDAAC/1196

```{r dendrometer2}
#| message: false
#| warning: false
read_csv("data/raw_data/LC14_Aboveground_Prod_1196/data/Dendrometer_DBH_REE_Plots_1999_2006.csv", 
         skip = 19, na = "-9999") %>% 
  rename_all(tolower) %>% 
  select(treatment, tree_id, life_form, family, genus_species, common_name,
         height_m, starts_with("dbh_"), date_death, field_notes) %>% 
  gather(date, dbh,
         -treatment, -tree_id, -life_form, -family, 
         -genus_species, -common_name,
         -height_m, -dbh_init,  -date_death, -field_notes) %>% 
  mutate(date = gsub("dbh_", "", date)) %>% 
  mutate(month = str_sub(date,1,3),
         year = str_sub(date,4,5)) %>% 
  mutate(date = paste0(year, "-", month, "-01")) %>% 
  mutate(date = as_date(date)) %>% 
  select(-year, -month) %>% 
  filter(life_form == "tree") %>% 
  ggplot(aes(date, dbh, 
             col = genus_species,
             group = paste(treatment, tree_id))) +
  geom_line(alpha = .2) +
  theme_bw() +
  xlab("") +
  facet_wrap(~ treatment) +
  ylab("Diameter at breast height [ cm ]") +
  scale_color_discrete(guide = "none") +
  scale_y_log10()
```

Can be used to assess TEE on dynamics simulation in TROLL.

```{r inventory}
#| message: false
#| warning: false
read_csv("data/raw_data/LC14_Aboveground_Prod_1196/data/Stem_inventory_REE_plots_2000_2004.csv", skip = 15, na = "-9999") %>% 
  rename_all(tolower) %>% 
  filter(life_form == "TREE") %>% 
  ggplot(aes(dbh_2000, fill = obs_final)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~ treatment, nrow = 2) +
  scale_x_log10() +
  scale_y_log10() +
  ylab("") + xlab("DBH 2000 [ cm ]") +
  scale_fill_discrete("")
```

To assess treatment effect on mortality and growth (DBH 2004 available).
