```{r set}
#| include: false
library(tidyverse)
```

# Fluxes {.unnumbered}

We have eddy flux data (evapotranspiration, gross primary productivity, PAR), soil water, litterfall, and leaf area index including control and TEE plots to evaluate and calibrate TROLL simulations.

*Data from eddy flux towers that are already shown in climate and that can be used for fluxes such as latent heat for evapotranspiration are not shown again here.*

*Midday and predawn potential are available to assess individual tree hydric stress and stomatal control in the model.*

## Undisturbed

### LBA-ECO CD-15 LAI and Productivity Data, km 67, Tapajos National Forest

Cite: Costa, M.H. and W. Cohen. 2013. LBA-ECO CD-15 LAI and Productivity Data, km 67, Tapajos National Forest: 2003-2004. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. http://dx.doi.org/10.3334/ORNLDAAC/1167

```{r lai}
#| message: false
#| warning: false
read_csv("data/raw_data/CD15_Productivity_1167/data/CD15_LAI_2004.csv",
  skip = 16, na = "-9999"
) %>%
  rename_all(tolower) %>%
  mutate(date = as_date(paste(year, month, day))) %>%
  select(-year, -month, -day, -n) %>%
  ggplot(aes(date, lai, col = as.character(plot))) +
  geom_line() +
  theme_bw() +
  scale_color_discrete(guide = "none") +
  xlab("") +
  ylab(expression("Leaf Area Index [" ~ m^2 ~ m^{
    -2
  } ~ "]"))
```

LAI data available for seasonal variation of canopy leaf dynamics assessment.

```{r litter}
#| message: false
#| warning: false
read_csv("data/raw_data/CD15_Productivity_1167/data/CD15_Litterfall_km67_2004.csv", # nolint
  skip = 10, na = "-9999"
) %>%
  rename_all(tolower) %>%
  rename(date = sample_date, litterfall = mass_litterfall) %>%
  mutate(date = paste0(
    str_sub(date, 1, 4), "-",
    str_sub(date, 5, 6), "-",
    str_sub(date, 7, 8)
  )) %>%
  mutate(date = as_date(date)) %>%
  ggplot(aes(date, litterfall, col = paste(plot, subplot))) +
  geom_line() +
  theme_bw() +
  scale_color_discrete(guide = "none") +
  xlab("") +
  ylab(expression("Litterfall [" ~ g ~ 0.25 ~ m^{
    -2
  } ~ "|time lapse]"))
```

Litterfall data available for seasonal variation of canopy productivity dynamics assessment.

### LBA-ECO CD-02 Leaf Water Potential, Forest and Pasture Sites, Para, Brazil: 2000-2001

Cite: Almeida, D.,T.F. Domingues, J. Ehleringer, L.A. Martinelli, C. Cook, L. Flanagan, J. Berry, and J.P. Ometto. 2012. LBA-ECO CD-02 Leaf Water Potential, Forest and Pasture Sites, Para, Brazil: 2000-2001. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. http://dx.doi.org/10.3334/ORNLDAAC/1100

```{r leaf_pot}
#| message: false
#| warning: false
read_csv("data/raw_data/CD02_Leaf_Water_Potential_1100/data/CD02_Leaf_Water_Potential.csv", skip = 22, na = "-9999") %>% # nolint
  rename_all(tolower) %>%
  filter(site == "STM km 67 tower") %>%
  mutate(date = paste0(
    str_sub(date, 1, 4), "-",
    str_sub(date, 5, 6), "-",
    str_sub(date, 7, 8)
  )) %>%
  mutate(date = as_date(date)) %>%
  select(date, time, scientific_name, starts_with("lwp")) %>%
  gather(measure, lwp, -date, -time, -scientific_name) %>%
  select(-measure) %>%
  ggplot(aes(date, lwp, col = scientific_name)) +
  geom_point() +
  facet_wrap(~time, nrow = 2, scales = "free_y") +
  theme_bw() +
  geom_smooth(se = FALSE) +
  scale_color_discrete("") +
  theme(legend.position = "bottom") +
  xlab("") +
  ylab("Leaf Water Potential [MPa]")
```

### LBA-ECO CD-10 FOREST LITTER DATA FOR KM 67 TOWER SITE, TAPAJOS NATIONAL FOREST

Cite: Rice, A.H., E. P. Hammond, S. R. Saleska, L. Hutyra, M. Palace, M. Keller, P. B. de Carmargo, K. Portilho, D. Marques and S. C. Wofsy. 2008. LBA-ECO CD-10 Forest Litter Data for km 67 Tower Site, Tapajos National Forest. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. doi:10.3334/ORNLDAAC/862

```{r litter2}
#| message: false
#| warning: false
read_csv("data/raw_data/CD10_Litter_Tapajos_862/data/lba_km67_litter_archive.txt", # nolint
  na = "-9999"
) %>%
  rename_all(tolower) %>%
  rename(date = yyyymmdd) %>%
  mutate(date = paste0(
    str_sub(date, 1, 4), "-",
    str_sub(date, 5, 6), "-",
    str_sub(date, 7, 8)
  )) %>%
  mutate(date = as_date(date)) %>%
  select(date, dt, leaf, fruit, wood, misc, total) %>%
  gather(type, litterfall, -date, -dt) %>%
  mutate(litterfall = as.numeric(litterfall)) %>%
  ggplot(aes(date, litterfall, col = type)) +
  geom_line() +
  theme_bw() +
  scale_color_discrete("") +
  xlab("") +
  ylab(expression("Litterfall [" ~ Mg ~ ha^{
    -2
  } ~ yr^{
    -1
  } ~ "]")) +
  scale_y_log10()
```

## TEE

### LBA-ECO CD-05 Soil VWC and Meteorology, Rainfall Exclusion, Tapajos National Forest

Cite: Nepstad, D.C., P.R. Moutinho, and P. Brando. 2013. LBA-ECO CD-05 Soil VWC and Meteorology, Rainfall Exclusion, Tapajos National Forest. Data set. Available on-line \[<http://daac.ornl.gov%5D> from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, USA <http://dx.doi.org/10.3334/ORNLDAAC/1169>

```{r vwc1}
#| message: false
#| warning: false
read_csv("data/raw_data/CD05_Micromet_1169/data/VWC_REE_plots_1999_2007.csv",
  skip = 15, na = "-9999"
) %>%
  rename_all(tolower) %>%
  mutate(date = paste0(
    str_sub(date, 1, 4), "-",
    str_sub(date, 5, 6), "-",
    str_sub(date, 7, 8)
  )) %>%
  mutate(date = as_datetime(date)) %>%
  ggplot(aes(date, vwc_reprocessed,
    group = paste(treatment, pit, side, depth), col = depth
  )) +
  geom_line() +
  facet_wrap(~treatment, nrow = 2) +
  theme_bw() +
  scale_color_viridis_c(expression("Depth [ m ]")) +
  xlab("") +
  ylab(expression("Volumetric Water Content [ VWC" ~ cm^3 ~ cm^{
    -3
  } ~ "]"))
```

```{r vwc2}
#| message: false
#| warning: false
read_csv("data/raw_data/CD05_Micromet_1169/data/VWC_REE_plots_1999_2007.csv",
  skip = 15, na = "-9999"
) %>%
  rename_all(tolower) %>%
  mutate(date = paste0(
    str_sub(date, 1, 4), "-",
    str_sub(date, 5, 6), "-",
    str_sub(date, 7, 8)
  )) %>%
  mutate(date = as_datetime(date)) %>%
  group_by(date, depth, treatment) %>%
  summarise(vwc = mean(vwc_reprocessed, na.rm = TRUE)) %>%
  ggplot(aes(date, depth, col = vwc)) +
  geom_point() +
  facet_wrap(~treatment, nrow = 2) +
  theme_bw() +
  scale_color_viridis_c(
    expression("Volumetric Water Content [ VWC" ~ cm^3 ~ cm^{
      -3
    } ~ "]"),
    direction = -1
  ) +
  xlab("") +
  ylab(expression("Depth [ m ]")) +
  theme(legend.position = "bottom")
```

To evaluated soil water dynamics.

### LBA-ECO ND-02 Soil Volumetric Water Content, Tapajos National Forest, Brazil

Cite: Belk, E.L., D. Markewitz, T. Rasmussen, E.J.M. Carvalho, D.C. Nepstad, and E.A. Davidson. 2012. LBA-ECO ND-02 Soil Volumetric Water Content, Tapajos National Forest, Brazil. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. http://dx.doi.org/10.3334/ORNLDAAC/1061

```{r vwc3}
#| message: false
#| warning: false
bind_rows(
  read_csv("data/raw_data/ND02_REE_Soil_VWC_1061/data/Seca_Floresta_actual_vs_simulated_VWC_control_plot.csv", # nolint
    skip = 36, na = "-9999"
  ),
  read_csv("data/raw_data/ND02_REE_Soil_VWC_1061/data/Seca_Floresta_actual_vs_simulated_VWC_treatment_plot.csv", # nolint
    skip = 36, na = "-9999"
  )
) %>%
  rename_all(tolower) %>%
  select(date, plot_type, starts_with("a")) %>%
  gather(depth, vwc, -date, -plot_type) %>%
  mutate(depth = as.numeric(gsub("a", "", depth))) %>%
  ggplot(aes(date, vwc,
    col = depth,
    group = paste(plot_type, depth)
  )) +
  geom_line() +
  theme_bw() +
  facet_wrap(~plot_type, nrow = 2) +
  scale_color_viridis_c(expression("Depth [ m ]")) +
  xlab("") +
  ylab(expression("Volumetric Water Content [ VWC" ~ cm^3 ~ cm^{
    -3
  } ~ "]"))
```

To evaluated soil water dynamics.

### LBA-ECO LC-14 Biophysical Measurements, Rainfall Exclusion, Tapajos National Forest

Cite: Nepstad, D.C. and P.R. Moutinho. 2013. LBA-ECO LC-14 Biophysical Measurements, Rainfall Exclusion, Tapajos National Forest. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, USA http://dx.doi.org/10.3334/ORNLDAAC/1196

```{r lai2}
#| message: false
#| warning: false
read_csv("data/raw_data/LC14_Aboveground_Prod_1196/data/LAI_REE_plots_1999_2006.csv", # nolint
  skip = 14, na = "-9999"
) %>%
  rename_all(tolower) %>%
  mutate(date = paste0(
    str_sub(date, 1, 4), "-",
    str_sub(date, 5, 6), "-",
    str_sub(date, 7, 8)
  )) %>%
  mutate(date = as_date(date)) %>%
  ggplot(aes(date, lai,
    group = paste(treatment, grid)
  )) +
  geom_line(alpha = .2) +
  theme_bw() +
  facet_wrap(~treatment, nrow = 2) +
  xlab("") +
  ylab(expression("Leaf Area Index [" ~ m^2 ~ m^{
    -2
  } ~ "]"))
```

LAI data available for seasonal variation of canopy leaf dynamics assessment.

```{r fine_litter}
#| message: false
#| warning: false
read_csv("data/raw_data/LC14_Aboveground_Prod_1196/data/Fine_Litter_REE_plots_1998_2006.csv", skip = 16, na = "-9999") %>% # nolint
  rename_all(tolower) %>%
  mutate(date = paste0(
    str_sub(date, 1, 4), "-",
    str_sub(date, 5, 6), "-",
    str_sub(date, 7, 8)
  )) %>%
  mutate(date = as_date(date)) %>%
  gather(type, mass, -treatment, -date, -point) %>%
  mutate(type = gsub("mass_", "", type)) %>%
  mutate(type = gsub("_", " ", type)) %>%
  na.omit() %>%
  ggplot(aes(date, mass,
    col = type,
    group = paste(treatment, point, type)
  )) +
  geom_line(alpha = 0.2) +
  facet_wrap(~treatment, nrow = 2) +
  theme_bw() +
  scale_y_log10() +
  xlab("") +
  ylab("Mass [" ~ g ~ m^{
    -2
  } ~ d^{
    -1
  } ~ "]")
```

Great data to assess TEE effect on fine litterfall.
