```{r set}
#| include: false
library(tidyverse)
```

# Fluxes {.unnumbered}

## Undisturbed

### LBA-ECO CD-15 LAI and Productivity Data, km 67, Tapajos National Forest

Cite: Costa, M.H. and W. Cohen. 2013. LBA-ECO CD-15 LAI and Productivity Data, km 67, Tapajos National Forest: 2003-2004. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. http://dx.doi.org/10.3334/ORNLDAAC/1167

```{r lai}
#| message: false
#| warning: false
read_csv("data/raw_data/CD15_Productivity_1167/data/CD15_LAI_2004.csv",
         skip = 16, na = "-9999") %>% 
  rename_all(tolower) %>% 
  mutate(date = as_date(paste(year, month, day))) %>% 
  select(-year, -month, -day, -n) %>% 
  ggplot(aes(date, lai, col = as.character(plot))) +
  geom_line() +
  theme_bw() +
  scale_color_discrete(guide = "none") +
  xlab("") +
  ylab(expression("Leaf Area Index ["~m^2~m^{-2}~"]"))
```

LAI data available for seasonal variation of canopy leaf dynamics assessment.

```{r litter}
#| message: false
#| warning: false
read_csv("data/raw_data/CD15_Productivity_1167/data/CD15_Litterfall_km67_2004.csv",
         skip = 10, na = "-9999") %>% 
  rename_all(tolower) %>% 
  rename(date = sample_date, litterfall = mass_litterfall) %>% 
  mutate(date = paste0(str_sub(date, 1,4), "-",
                       str_sub(date, 5,6), "-",
                       str_sub(date, 7,8))) %>% 
  mutate(date = as_date(date)) %>% 
  ggplot(aes(date, litterfall, col = paste(plot, subplot))) +
  geom_line() +
  theme_bw() +
  scale_color_discrete(guide = "none") +
  xlab("") +
  ylab(expression("Litterfall ["~g~0.25~m^{-2}~"|time lapse]"))
```

Litterfall data available for seasonal variation of canopy productivity dynamics assessment.

## TEE

### LBA-ECO CD-05 Soil VWC and Meteorology, Rainfall Exclusion, Tapajos National Forest

Cite: Nepstad, D.C., P.R. Moutinho, and P. Brando. 2013. LBA-ECO CD-05 Soil VWC and Meteorology, Rainfall Exclusion, Tapajos National Forest. Data set. Available on-line \[[http://daac.ornl.gov\]](http://daac.ornl.gov]) from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, USA <http://dx.doi.org/10.3334/ORNLDAAC/1169>

```{r vwc1}
#| message: false
#| warning: false
read_csv("data/raw_data/CD05_Micromet_1169/data/VWC_REE_plots_1999_2007.csv",
         skip = 15, na = "-9999") %>% 
  rename_all(tolower) %>% 
  mutate(date = paste0(str_sub(date, 1,4), "-",
                       str_sub(date, 5,6), "-",
                       str_sub(date, 7,8))) %>% 
  mutate(date = as_datetime(date)) %>% 
  ggplot(aes(date, vwc_reprocessed, 
             group = paste(treatment, pit, side, depth), col = depth)) +
  geom_line() +
  facet_wrap(~ treatment, nrow = 2) +
  theme_bw() +
  scale_color_viridis_c(expression("Depth [ m ]")) +
  xlab("") + ylab(expression("Volumetric Water Content [ VWC"~cm^3~cm^{-3}~"]"))
```

```{r vwc2}
#| message: false
#| warning: false
read_csv("data/raw_data/CD05_Micromet_1169/data/VWC_REE_plots_1999_2007.csv",
         skip = 15, na = "-9999") %>% 
  rename_all(tolower) %>% 
  mutate(date = paste0(str_sub(date, 1,4), "-",
                       str_sub(date, 5,6), "-",
                       str_sub(date, 7,8))) %>% 
  mutate(date = as_datetime(date)) %>% 
  group_by(date, depth, treatment) %>% 
  summarise(vwc = mean(vwc_reprocessed, na.rm = TRUE)) %>% 
  ggplot(aes(date, depth, col = vwc)) +
  geom_point() +
  facet_wrap(~ treatment, nrow = 2) +
  theme_bw() +
  scale_color_viridis_c(expression("Volumetric Water Content [ VWC"~cm^3~cm^{-3}~"]"),
                        direction = -1) +
  xlab("") + ylab(expression("Depth [ m ]")) +
  theme(legend.position = "bottom")
```

To evaluated soil water dynamics.

### LBA-ECO ND-02 Soil Volumetric Water Content, Tapajos National Forest, Brazil

Cite: Belk, E.L., D. Markewitz, T. Rasmussen, E.J.M. Carvalho, D.C. Nepstad, and E.A. Davidson. 2012. LBA-ECO ND-02 Soil Volumetric Water Content, Tapajos National Forest, Brazil. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, U.S.A. http://dx.doi.org/10.3334/ORNLDAAC/1061

```{r vwc3}
#| message: false
#| warning: false
bind_rows(
  read_csv("data/raw_data/ND02_REE_Soil_VWC_1061/data/Seca_Floresta_actual_vs_simulated_VWC_control_plot.csv",
         skip = 36, na = "-9999"),
  read_csv("data/raw_data/ND02_REE_Soil_VWC_1061/data/Seca_Floresta_actual_vs_simulated_VWC_treatment_plot.csv",
         skip = 36, na = "-9999")) %>% 
  rename_all(tolower) %>% 
  select(date, plot_type, starts_with("a")) %>% 
  gather(depth, vwc, -date, -plot_type) %>% 
  mutate(depth = as.numeric(gsub("a", "", depth))) %>% 
  ggplot(aes(date, vwc, col = depth, 
             group = paste(plot_type, depth))) +
  geom_line() +
  theme_bw() +
  facet_wrap(~ plot_type, nrow = 2) +
  scale_color_viridis_c(expression("Depth [ m ]")) +
  xlab("") + ylab(expression("Volumetric Water Content [ VWC"~cm^3~cm^{-3}~"]"))
```

To evaluated soil water dynamics.

### LBA-ECO LC-14 Biophysical Measurements, Rainfall Exclusion, Tapajos National Forest

Cite: Nepstad, D.C. and P.R. Moutinho. 2013. LBA-ECO LC-14 Biophysical Measurements, Rainfall Exclusion, Tapajos National Forest. Data set. Available on-line \[http://daac.ornl.gov\] from Oak Ridge National Laboratory Distributed Active Archive Center, Oak Ridge, Tennessee, USA http://dx.doi.org/10.3334/ORNLDAAC/1196

```{r lai2}
#| message: false
#| warning: false
read_csv("data/raw_data/LC14_Aboveground_Prod_1196/data/LAI_REE_plots_1999_2006.csv",
         skip = 14, na = "-9999") %>% 
  rename_all(tolower) %>% 
   mutate(date = paste0(str_sub(date, 1,4), "-",
                       str_sub(date, 5,6), "-",
                       str_sub(date, 7,8))) %>% 
  mutate(date = as_date(date)) %>% 
  ggplot(aes(date, lai, 
             group = paste(treatment, grid))) +
  geom_line(alpha = .2) +
  theme_bw() +
  facet_wrap(~ treatment, nrow = 2) +
  xlab("") +
  ylab(expression("Leaf Area Index ["~m^2~m^{-2}~"]"))
```

LAI data available for seasonal variation of canopy leaf dynamics assessment.

```{r fine_litter}
#| message: false
#| warning: false
read_csv("data/raw_data/LC14_Aboveground_Prod_1196/data/Fine_Litter_REE_plots_1998_2006.csv", skip = 16, na = "-9999") %>% 
  rename_all(tolower) %>% 
  mutate(date = paste0(str_sub(date, 1,4), "-",
                       str_sub(date, 5,6), "-",
                       str_sub(date, 7,8))) %>% 
  mutate(date = as_date(date)) %>% 
  gather(type, mass, -treatment, -date, -point) %>% 
  mutate(type = gsub("mass_", "", type)) %>% 
  mutate(type = gsub("_", " ", type)) %>% 
  na.omit() %>% 
  ggplot(aes(date, mass, col = type, 
             group = paste(treatment, point, type))) +
  geom_line(alpha = 0.2) +
  facet_wrap(~ treatment, nrow = 2) +
  theme_bw() +
  scale_y_log10() +
  xlab("") +
  ylab("Mass ["~g~m^{-2}~d^{-1}~"]")
```

Great data to assess TEE effect on fine litterfall.
