```{r set}
#| include: false
library(tidyverse)
```

# Future climate {.unnumbered}

> **Summary.** Future climate forcing from CMIP6 GCM between 1980 and 2100 under low emission scenarii SSP1-2.6 and high emission scenarii SSP5-8.5 [@neill2016]. Daily values were statistically downscaled on ERA5-Land historical data per month using cumulative distribution function transformation CDFt [@vrac2015]. Half-hourly values were downscaled using daily mean variations across years from ERA5-Land.

```{r assemble}
#| eval: false
library(ncdf4)
get_var <- function(file, var) {
  nc <- try(nc_open(file))
  if (!inherits(nc, "try-error")) {
    data <- tibble(
      gcm = ncatt_get(nc, 0, "parent_source_id")$value,
      experiment = ncatt_get(nc, 0, "experiment_id")$value,
      date = as.character(as_date(ncatt_get(nc, "time", "units")$value) +
        ncvar_get(nc, "time")),
      "{var}" := as.numeric(ncvar_get(nc, var))
    )
    nc_close(nc)
    data
  } else {
    warning(paste("Failed:", file))
    tibble()
  }
}
data <- lapply(c("pr", "tas", "sfcWind", "hurs", "rsus", "rsds"),
             function(var)
               list.files(paste0("data/raw_data/cmip6/",
                                 var, "/"), full.names = TRUE) %>%
               lapply(get_var, var) %>%
               bind_rows()) %>% 
  reduce(full_join)
t <- data %>% 
  mutate(pr = pr * 60 * 60 * 24) %>% 
  mutate(tas = tas - 273.15) %>% 
  rename(ws = sfcWind) %>% 
  mutate(snet = rsds - rsus) %>% 
  select(-rsds, -rsus) %>% 
  mutate(vpd = 0.6112*exp((17.62*tas)/(243.12+tas))*(1-hurs/100)) %>% 
  select(-hurs) %>% 
  filter(paste(month(date), day(date)) != "2 29") %>% 
  filter(!(gcm %in% c("CanESM5", "ACCESS-CM2")))
# removing uncomplete GCMs for the moment, no better plan yet
t %>% 
  gather(variable, value, -gcm, -experiment, -date) %>% 
  filter(is.na(value)) %>% 
  select(gcm, experiment, variable) %>% 
  unique() %>% 
  mutate(value = 1) %>% 
  pivot_wider(names_from = variable, values_from = value)
t %>% 
  arrange(gcm, experiment, date) %>%
  write_tsv("data/derived_data/projections_raw.tsv")
```

```{r bias-correct}
#| eval: false
era <- read_tsv("data/derived_data/climate.tsv") %>%
  gather(variable, value, -date) %>%
  na.omit() %>%
  group_by(variable, date = as_date(date)) %>%
  mutate(value = ifelse(variable %in% c("pr"), sum(value), value)) %>%
  summarise(era = mean(value)) %>%
  ungroup()
proj <- read_tsv("data/derived_data/projections_raw.tsv") %>%
  gather(variable, value, -date, -gcm, -experiment)
apply_bc <- function(model, month, var, era, proj) {
  era_sub <- era %>%
    filter(variable == var, month(date) == month)
  proj_sub <- proj %>%
    filter(gcm == model, variable == var, month(date) == month) %>%
    left_join(era_sub, by = join_by(date, variable))
  train <- filter(proj_sub, !is.na(era))
  adjusted <- CDFt::CDFt(
    ObsRp = train$era,
    DataGp = train$value,
    DataGf = proj_sub$value
  )
  print(paste(model, month, var, "done."))
  proj_sub %>%
    mutate(value = adjusted$DS) %>%
    select(-era)
}
bc <- expand_grid(
  gcm = unique(read_tsv("data/derived_data/projections_raw.tsv")$gcm),
  month = 1:12,
  var = c("pr", "tas", "ws", "snet", "vpd")
) %>%
  rowwise() %>%
  mutate(bc = list(apply_bc(gcm, month, var, era, proj)))
bc %>%
  unnest() %>%
  select(-gcm1, -var, -month) %>%
  pivot_wider(names_from = variable, values_from = value) %>%
  arrange(gcm, experiment, date) %>%
  write_tsv("data/derived_data/projections.tsv")
```

## Precipitation

CMIP6 raw precipitation projections mostly underestimated annual precipitation with lower monthly values and unrealistic seasonal pattern showing lagged and more intense dry seasons for all GCMs. After bias-correction we obtained more realistic annual precipitation, good monthly precipitation correlations to ERA5-Land above 0.6 and a realistic seasonality with the June to November dry season below 100-mm. Bias-correction using cumulative distribution function was applied on absolute precipitation value but might be improved considering relative daily precipitation volume in the month.

```{r pr_an_raw}
#| message: false
#| warning: false
#| fig-cap: "Annual CMIP6 raw precipitation projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "year"))) %>%
  summarise(pr = sum(pr, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land", experiment = "historical")
read_tsv("data/derived_data/projections_raw.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "year")) %>%
  summarise(pr = sum(pr, na.rm = TRUE)) %>%
  ggplot(aes(date, pr, group = paste(gcm, experiment), col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab("Precipitation [ mm ]") +
  scale_color_discrete("")
```

```{r pr_mon_raw}
#| message: false
#| warning: false
#| fig-cap: "Evaluations of CMIP6 raw projections monthly precipitations against ERA5-Land." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "month"))) %>%
  summarise(era = sum(pr, na.rm = TRUE))
read_tsv("data/derived_data/projections_raw.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "month")) %>%
  summarise(pr = sum(pr, na.rm = TRUE)) %>%
  left_join(era) %>%
  ggplot(aes(era, pr, col = gcm)) +
  geom_abline() +
  geom_point(alpha = .2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  ggpubr::stat_cor() +
  scale_color_discrete("") +
  ggtitle("Monthly total precipitation [ mm ]") +
  xlab("ERA5-Land") +
  ylab("CMIP6 GCM raw projection")
```

```{r pr_seas_raw}
#| message: false
#| warning: false
#| fig-cap: "Seasonal CMIP6 raw precipitation projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "month"))) %>%
  summarise(pr = sum(pr, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land") %>%
  group_by(gcm, date = month(date)) %>%
  summarise(pr = mean(pr))
read_tsv("data/derived_data/projections_raw.tsv") %>%
  filter(experiment == "historical") %>%
  group_by(gcm, date = floor_date(date, "month")) %>%
  summarise(pr = sum(pr, na.rm = TRUE)) %>%
  group_by(gcm, date = month(date)) %>%
  summarise(pr = mean(pr)) %>%
  ggplot(aes(date, pr, col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab("Precipitation [ mm ]") +
  scale_color_discrete("") +
  scale_x_continuous(
    breaks = 1:12,
    labels = c(
      "J", "F", "M", "A", "M", "J", "J", "A",
      "S", "O", "N", "D"
    )
  )
```

```{r pr_an_bc}
#| message: false
#| warning: false
#| fig-cap: "Annual CMIP6 bias-corrected precipitation projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "year"))) %>%
  summarise(pr = sum(pr, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land", experiment = "historical")
read_tsv("data/derived_data/projections.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "year")) %>%
  summarise(pr = sum(pr)) %>%
  ggplot(aes(date, pr, group = paste(gcm, experiment), col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab("Precipitation [ mm ]") +
  scale_color_discrete("")
```

```{r pr_mon_bc}
#| message: false
#| warning: false
#| fig-cap: "Evaluations of CMIP6 bias-corrected projections monthly precipitations against ERA5-Land." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "month"))) %>%
  summarise(era = sum(pr, na.rm = TRUE))
read_tsv("data/derived_data/projections.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "month")) %>%
  summarise(pr = sum(pr)) %>%
  left_join(era) %>%
  ggplot(aes(era, pr, col = gcm)) +
  geom_abline() +
  geom_point(alpha = .2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  ggpubr::stat_cor() +
  scale_color_discrete("") +
  ggtitle("Monthly total precipitation [ mm ]") +
  xlab("ERA5-Land") +
  ylab("CMIP6 GCM raw projection")
```

```{r pr_seas_bc}
#| message: false
#| warning: false
#| fig-cap: "Seasonal CMIP6 bias-corrected precipitation projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "month"))) %>%
  summarise(pr = sum(pr, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land") %>%
  group_by(gcm, date = month(date)) %>%
  summarise(pr = mean(pr))
read_tsv("data/derived_data/projections.tsv") %>%
  filter(experiment == "historical") %>%
  group_by(gcm, date = floor_date(date, "month")) %>%
  summarise(pr = sum(pr)) %>%
  group_by(gcm, date = month(date)) %>%
  summarise(pr = mean(pr)) %>%
  ggplot(aes(date, pr, col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab("Precipitation [ mm ]") +
  scale_color_discrete("") +
  scale_x_continuous(
    breaks = 1:12,
    labels = c(
      "J", "F", "M", "A", "M", "J", "J", "A",
      "S", "O", "N", "D"
    )
  )
```

## Temperature

CMIP6 raw temperature projections were already pretty good but showed better mean annual pattern after bias correction still showing a bit of underestimations of temperature for month with high temperature. The seasonal pattern was good after bias correction.

```{r tas_an_raw}
#| message: false
#| warning: false
#| fig-cap: "Annual CMIP6 mean temperature projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "year"))) %>%
  summarise(tas = mean(tas, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land", experiment = "historical")
read_tsv("data/derived_data/projections_raw.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "year")) %>%
  summarise(tas = mean(tas, na.rm = TRUE)) %>%
  ggplot(aes(date, tas, group = paste(gcm, experiment), col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab("Temperature [ °C ]") +
  scale_color_discrete("")
```

```{r tas_an_bc}
#| message: false
#| warning: false
#| fig-cap: "Annual CMIP6 bias-corrected mean temperature projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "year"))) %>%
  summarise(tas = mean(tas, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land", experiment = "historical")
read_tsv("data/derived_data/projections.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "year")) %>%
  summarise(tas = mean(tas, na.rm = TRUE)) %>%
  ggplot(aes(date, tas, group = paste(gcm, experiment), col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab("Temperature [ °C ]") +
  scale_color_discrete("")
```

```{r tas_mon_bc}
#| message: false
#| warning: false
#| fig-cap: "Evaluations of CMIP6 bias-corrected projections monthly mean temperature against ERA5-Land." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "month"))) %>%
  summarise(era = mean(tas, na.rm = TRUE))
read_tsv("data/derived_data/projections.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "month")) %>%
  summarise(tas = mean(tas)) %>%
  left_join(era) %>%
  ggplot(aes(era, tas, col = gcm)) +
  geom_abline() +
  geom_point(alpha = .2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  ggpubr::stat_cor() +
  scale_color_discrete("") +
  ggtitle("Monthly mean temperature [ °C ]") +
  xlab("ERA5-Land") +
  ylab("CMIP6 GCM raw projection")
```

```{r tas_seas_bc}
#| message: false
#| warning: false
#| fig-cap: "Seasonal CMIP6 bias-corrected mean temperature projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "month"))) %>%
  summarise(tas = mean(tas, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land") %>%
  group_by(gcm, date = month(date)) %>%
  summarise(tas = mean(tas))
read_tsv("data/derived_data/projections.tsv") %>%
  filter(experiment == "historical") %>%
  group_by(gcm, date = floor_date(date, "month")) %>%
  summarise(tas = mean(tas)) %>%
  group_by(gcm, date = month(date)) %>%
  summarise(tas = mean(tas)) %>%
  ggplot(aes(date, tas, col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab("Precipitation [ mm ]") +
  scale_color_discrete("") +
  scale_x_continuous(
    breaks = 1:12,
    labels = c(
      "J", "F", "M", "A", "M", "J", "J", "A",
      "S", "O", "N", "D"
    )
  )
```

## Wind speed

CMIP6 raw wind speed showed exaggerated wind speed with 2 to 3 folds, fixed by bias correction.

```{r ws_an_raw}
#| message: false
#| warning: false
#| fig-cap: "Annual CMIP6 mean wind speed projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "year"))) %>%
  summarise(ws = mean(ws, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land", experiment = "historical")
read_tsv("data/derived_data/projections_raw.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "year")) %>%
  summarise(ws = mean(ws, na.rm = TRUE)) %>%
  ggplot(aes(date, ws, group = paste(gcm, experiment), col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab(expression("Wind speed ["~m~s^{-1}~"]")) +
  scale_color_discrete("")
```

```{r ws_an_bc}
#| message: false
#| warning: false
#| fig-cap: "Annual CMIP6 bias-corrected mean wind speed projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "year"))) %>%
  summarise(ws = mean(ws, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land", experiment = "historical")
read_tsv("data/derived_data/projections.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "year")) %>%
  summarise(ws = mean(ws, na.rm = TRUE)) %>%
  ggplot(aes(date, ws, group = paste(gcm, experiment), col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab(expression("Wind speed ["~m~s^{-1}~"]")) +
  scale_color_discrete("")
```

## Vapour pressure deficit

Vapour pressure deficit was derived from CMIP6 raw near surface relative humidity hurs and temperature at surface tas with the following formula:

$$
vpd = 0.6112 \times e^{\frac{17.62 \times tas}{243.12+tas}} \times (1-hurs)
$$

CMIP6 raw vapour pressure deficit showed exaggerated deficit, fixed by bias correction.

```{r vpd_an_raw}
#| message: false
#| warning: false
#| fig-cap: "Annual CMIP6 mean vpd projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "year"))) %>%
  summarise(vpd = mean(vpd, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land", experiment = "historical")
read_tsv("data/derived_data/projections_raw.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "year")) %>%
  summarise(vpd = mean(vpd, na.rm = TRUE)) %>%
  ggplot(aes(date, vpd, group = paste(gcm, experiment), col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab(expression("Vapour Pressure Deficit ["~Pa~"]")) +
  scale_color_discrete("")
```

```{r vpd_an_bc}
#| message: false
#| warning: false
#| fig-cap: "Annual CMIP6 bias-corrected mean vpd projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "year"))) %>%
  summarise(vpd = mean(vpd, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land", experiment = "historical")
read_tsv("data/derived_data/projections.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "year")) %>%
  summarise(vpd = mean(vpd, na.rm = TRUE)) %>%
  ggplot(aes(date, vpd, group = paste(gcm, experiment), col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab(expression("Vapour Pressure Deficit ["~Pa~"]")) +
  scale_color_discrete("")
```

## Radiations

Net solar raditions snet was derived from CMIP6 raw surface downwelling and upwelling shortwave radiations rsdus and rsus with the following formula:

$$
snet = rsdus - rsus
$$

CMIP6 raw solar radiations showed exaggerated radiations, fixed by bias correction.

```{r snet_an_raw}
#| message: false
#| warning: false
#| fig-cap: "Annual CMIP6 mean snet projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "year"))) %>%
  summarise(snet = mean(snet, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land", experiment = "historical")
read_tsv("data/derived_data/projections_raw.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "year")) %>%
  summarise(snet = mean(snet, na.rm = TRUE)) %>%
  ggplot(aes(date, snet, group = paste(gcm, experiment), col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab(expression("Net radiations ["~W~m^{-2}~"]")) +
  scale_color_discrete("")
```

```{r snet_an_bc}
#| message: false
#| warning: false
#| fig-cap: "Annual CMIP6 bias-corrected mean snet projections for scenarii SSP1-2.6 and SSP5-8.5 compared to ERA5-Land reanalysis (black)." #nolint
era <- read_tsv("data/derived_data/climate.tsv") %>%
  group_by(date = as_date(floor_date(date, "year"))) %>%
  summarise(snet = mean(snet, na.rm = TRUE)) %>%
  mutate(gcm = "ERA5-Land", experiment = "historical")
read_tsv("data/derived_data/projections.tsv") %>%
  group_by(gcm, experiment, date = floor_date(date, "year")) %>%
  summarise(snet = mean(snet, na.rm = TRUE)) %>%
  ggplot(aes(date, snet, group = paste(gcm, experiment), col = gcm)) +
  geom_line() +
  geom_line(data = era, col = "black") +
  theme_bw() +
  xlab("") +
  ylab(expression("Net radiations ["~W~m^{-2}~"]")) +
  scale_color_discrete("")
```

## Half-hourly downscaling

*ToDo using each day half-hourly mean variation across years.*

```{r hh_var}
#| message: false
#| warning: false
#| fig-cap: "Half-hourly relative variation across 45 years from the first day of the month" #nolint
read_tsv("data/derived_data/climate.tsv") %>%
  select(-pr) %>%
  gather(variable, value, -date) %>%
  filter(day(date) == 1) %>%
  mutate(
    time = minute(date) + 60 * hour(date),
    month = month(date)
  ) %>%
  group_by(variable, date2 = as_date(date)) %>%
  mutate(value = value / mean(value)) %>%
  select(-date2) %>%
  group_by(variable, month, time) %>%
  summarise(
    l = quantile(value, .05),
    m = quantile(value, .5),
    h = quantile(value, .95)
  ) %>%
  ggplot(aes(time / 60, m)) +
  geom_ribbon(aes(ymin = l, ymax = h), fill = "lightgrey", col = NA) +
  geom_line() +
  facet_grid(variable ~ month, scales = "free") +
  theme_bw() +
  xlab("") +
  ylab("") +
  theme(axis.text.x = element_text(angle = 90))
```

```{r hh_var_apply}
#| eval: false
library(tidyverse)
hh_var <- read_tsv("data/derived_data/climate.tsv") %>%
  select(-pr) %>% 
  gather(variable, value, -date) %>%
  mutate(
    year = year(date),
    month = month(date),
    day = day(date),
    time = minute(date) + 60 * hour(date)
  ) %>% 
  group_by(variable, year, month, day) %>%
  mutate(realtive_variation = value / mean(value, na.rm = TRUE)) %>% 
  select(-year) %>% 
  group_by(variable, month, day, time) %>%
  summarise(realtive_variation = mean(realtive_variation, na.rm = TRUE))
hh_var <- hh_var %>% 
  bind_rows(filter(hh_var, variable == "snet", time == 0) %>% 
              mutate(variable = "pr") %>% 
              mutate(realtive_variation = 1))
t <- read_tsv("data/derived_data/projections.tsv") %>% 
  gather(variable, value, -gcm, -experiment, -date) %>% 
  mutate(
    month = month(date),
    day = day(date)
  ) %>% 
  left_join(hh_var, relationship = "many-to-many") %>% 
  mutate(value = value * realtive_variation) %>% 
  mutate(date = as_datetime(date) + time*60) %>% 
  select(-month, -day, -time, -realtive_variation) %>% 
  pivot_wider(names_from = variable, values_from = value)
t %>% 
  arrange(gcm, experiment, date) %>%
  write_tsv("data/derived_data/projections_hh.tsv")
```

```{r hh_proj}
#| message: false
#| warning: false
#| fig-cap: "CMIP6 bias-corrected and half-hourly downscaled monthly means from 1980 to 2025."
read_tsv("data/derived_data/projections_hh.tsv") %>% 
  filter(experiment == "historical") %>%
  select(-experiment) %>% 
  filter(year(date) %in% 1980:2024) %>% 
  gather(variable, value, -date, -gcm) %>% 
  group_by(gcm, variable, date = floor_date(date, "month")) %>%
  mutate(value = ifelse(variable %in% c("pr"),
    sum(value, na.rm = TRUE), value
  )) %>%
  summarise_all(mean) %>%
  ggplot(aes(date, value, col = gcm)) +
  geom_line() +
  facet_wrap(~variable, scales = "free_y", nrow = 5) +
  theme_bw() +
  xlab("") +
  ylab("") +
  scale_color_discrete(guide = "none")
```
