```{r set}
#| include: false
library(tidyverse)
```

# Spin-up {.unnumbered}

The first spin-up without calibration (using the one of @schmitt2024 based on @rice2004 ) gave acceptable but sub-optimal forest structure. The biomass is slightly under the one given by @nepstad2002, @powell2013 and @rice2004. The basal area is strongly underestimated according to @nepstad2002. And the number of stems is good according @rice2004, but logically as it is the data used for calibration. @rice2004 study site and inventory protocol differs from TEE and @nepstad2002. Thus it is possible that the calibration does not represent well the control forest from TEE. Nonetheless, we will proceed for the moment with this spin-up to develop the proof of concept, looking at relative hcange in biomass [@powell2013] and not absolute value.

```{r spinup}
#| message: false
#| warning: false
#| fig-cap: "Caption."
read_tsv("simulations/results0/spinup/R1/R1_0_sumstats.txt") %>% 
  mutate(date = as_date("0000-01-01") + iter) %>% 
  select(date, sum10, ba10, agb) %>% 
  gather(variable, value, -date) %>% 
  ggplot(aes(date, value)) +
  geom_line() +
  facet_wrap(~ variable, nrow = 3, scales = "free_y") +
  theme_bw() +
  geom_hline(aes(yintercept = value, col = "Nepstad et al. (2002)"),
             data_frame(value = c(305*10^3, 30),
                        variable = c("agb", "ba10"))) +
  geom_hline(aes(yintercept = value, col = "Rice et al. (2004)"),
             data_frame(value = c(470, 2*143.7*10^3),
                        variable = c("sum10", "agb"))) +
  geom_hline(aes(yintercept = value, col = "Powell et al. (2013)"),
             data_frame(value = c(2*144*10^3),
                        variable = c("agb"))) +
  scale_color_discrete("") +
  theme(legend.position = "bottom", axis.title = element_blank())
```
